---
title: "PolyBailout_Script"
author: "Pooja Pednekar"
date: "2026-05-22"
output: html_document
---
As of 05/22/2026 this script plots pH, Temperature and Conductivity data from the APEX loggers. This script also plots PreSens Oxy data which is used for the first time to include in Polyp bailout experiments. Additionally, there is a PAM plot included in this script which gives us the background on the photosynthetic ability of the corals. 

Probe #s 1 and 2 are probes placed in the control chambers of the experiment while probes 3 and 4 are placed in the experimented chambers. We have two treatment groups: Control (Normal Salinity) and Experimental (High Salinity). Even though we have 4 replicates per treatment group, we only have been so far measuring data in two replicates per treatment group for all the variables. 

This script plots Polyp Bailout measurements for the Putnam lab and is based on the Putnam Lab Script available here:  <https://github.com/Putnam-Lab/CBLS_Wetlab/tree/main/Scripts/daily-measurements.Rmd>

For APEX data, the script is based on Zoe Dellart's Protocol<https://github.com/zdellaert/TimeSeries/blob/main/protocols/Apex_Download.md> and script <https://github.com/zdellaert/TimeSeries/blob/main/3-Pacu/code/DMs.Rmd>.

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate) # used for converting 8 digit date into datetime format for R
library(RColorBrewer)
library(rmarkdown)
library(tinytex)
library(xml2)
library(plyr)
library(dplyr)
library(ggpmisc)
library(gsw)
library(janitor)
library(ggpubr)
library(lme4)
library(lmerTest)
library(emmeans)

## If seacarb needs to be downloaded:
#packageurl <- "https://cran.r-project.org/src/contrib/Archive/seacarb/seacarb_3.2.tar.gz"
#install.packages(packageurl, repos = NULL, type = "source")
#install.packages("seacarb")
library(seacarb) 


custom_colors <- c("Tmp_1" = "#9ECAE1", "Cond_1" = "#9ECAE1", "pH_1" = "#9ECAE1", "Tmp_2" = "#3182BD", "Cond_2" = "#3182BD", "pH_2" = "#3182BD", "Tmp_3" = "#FC9272", "Cond_3" = "#FC9272", "pH_3" = "#FC9272", "Tmp_4" = "#E15759", "Cond_4" = "#E15759", "pH_4" = "#E15759")
```

##APEX DATA conversion 
```{r}
#exported apex data online
#1. Export data
#2.Open xml file in excel
#3. Save as csv
#4. Importing csv into R


#reading file
apex <- read.csv("C:/PUTNAM LAB/Pacuta_Polyp_Bailout/data/ApexLogger/Spring_2026/20260422Trail1/ApexRawData.csv", sep=",", skip=c(1), header=TRUE, na.strings = "NA")[6:9]

#Selecting column names of our desire
colnames(apex) <-  c("DateTime","ProbeName","ProbeType","Value")

head(apex)

#converting date and adapting to the timezone
apex$DateTimeHST <- parse_date_time(apex$DateTime, "%m/%d/%y %H:%M:%S", tz = "Pacific/Honolulu")
apex$DateTimeEST <- with_tz(apex$DateTimeHST, tzone = "America/New_York")
head(apex)

# filtering dates based on experiment timeline  (use GMT for the time, so +6)
apex_Exp <- apex %>% filter(DateTimeEST >= "2026-04-22 11:00:00" & DateTimeEST <= "2026-04-23 11:00:00")

#unique(apex_Exp$ProbeType) #checking the probetype names

1.00#filtering data to be plotted  filter(!(ProbeType == "Cond" & Value %in% c(0, 0.1)))

apex_toplot <- apex_Exp %>%
  filter(ProbeType == "Temp"|
         ProbeType == "pH"|
         ProbeType == "Cond") %>%
  filter(!(ProbeType == "Cond" & Value <= 0.2)) %>%
  mutate(Value = ifelse(ProbeType == "Cond",round(Value * 0.64, 1), Value))
```
## APEX plots
```{r}
# Threshold lines
hlines_data <- data.frame(yintercept = c(22.0, 30.0,35.0, 45.0,8.1, 8.3),
  ProbeType = c("Temp", "Temp", "Cond", "Cond","pH", "pH"))

facet_labels <- unique(apex_Exp$ProbeType)
names(facet_labels) = facet_labels
facet_labels <- replace(facet_labels, which(facet_labels == "pH"), "pH")
facet_labels <- replace(facet_labels, which(facet_labels == "Cond"), "Salinity (psu)")
facet_labels <- replace(facet_labels, which(facet_labels == "Temp"), "Temperature (ºC)")

apex_final <- apex_toplot %>%
  ggplot(aes(x = DateTimeEST,
             y = Value,
             colour = ProbeName)) +
  geom_smooth(size = 0.8, alpha = 0.7) +
  geom_hline(data = hlines_data,aes(yintercept = yintercept),linetype = "dashed",
    inherit.aes = FALSE
  ) +
  scale_color_manual(values = custom_colors) +
  xlab("DateTimeEST") +
  facet_grid(factor(ProbeType, c("pH", "Cond", "Temp")) ~ .,scales = "free_y",labeller = as_labeller(facet_labels)) +
  theme_bw() +
  theme(text = element_text(size = 14))

# Save plot 
ggsave("../Output/Daily_Measurements.png", apex_final, width = 20, height = 15, units = c("in"))

apex_final
```



#OXY DATA
```{r}
#reading oxy files
Oxy1 <- read.csv("C:/PUTNAM LAB/Pacuta_Polyp_Bailout/data/PreSens_OxyData/20260422/20260422_B2_CH1.csv",
  fileEncoding = "latin1")
Oxy2 <- read.csv("C:/PUTNAM LAB/Pacuta_Polyp_Bailout/data/PreSens_OxyData/20260422/20260422_B4_CH2.csv",
  fileEncoding = "latin1")
Oxy3 <- read.csv("C:/PUTNAM LAB/Pacuta_Polyp_Bailout/data/PreSens_OxyData/20260422/20260422_A2_CH3.csv",
  fileEncoding = "latin1")
Oxy4 <- read.csv("C:/PUTNAM LAB/Pacuta_Polyp_Bailout/data/PreSens_OxyData/20260422/20260422_A4_CH5.csv",
  fileEncoding = "latin1")


#adding columns in the dataframes
Oxy1$Chamber <- "Oxy_1"
Oxy2$Chamber <- "Oxy_2"
Oxy3$Chamber <- "Oxy_3"
Oxy4$Chamber <- "Oxy_4"

# Combine into one dataframe
oxy_everything <- bind_rows(Oxy1, Oxy2, Oxy3, Oxy4)

#select columns needed for the analysis 
oxy_final <- oxy_everything %>%
  select(Date, Oxygen, Chamber)%>%
 filter(!(Chamber == "Oxy_2" & Oxygen > 100))

# Convert datetime column
oxy_final$Date <- as.POSIXct(
  oxy_final$Date,
  format = "%d/%m/%y %H:%M:%S"
)

# Plot
oxy_plot <- ggplot(oxy_final,
       aes(x = Date,
           y = Oxygen,
           color = Chamber)) +
  geom_line(alpha = 0.7) +
  theme_classic() +
  labs(x = "Time",
       y = "Oxygen Concentration (µmol/L)",
       color = "Chamber")


# Save plot 
ggsave("../Output/20260422/Oxygen_Trial1.png", oxy_plot, width = 10, height = 8, units = c("in"))


```



#PAM DATA
```{r}

PAM <- read.csv("C:/PUTNAM LAB/Pacuta_Polyp_Bailout/data/PAM/PAM.csv") %>% clean_names()
PAM <- PAM %>% mutate(date = as.factor(date)) %>% 
                mutate(timepoint = factor(timepoint)) %>%
                mutate(treatment = factor(treatment,levels = c("A1","A2","A3","A4","B1", "B2", "B3", "B4" , ordered=TRUE))) %>%
                dplyr::rename("replicate" = treatment)%>%
                mutate(treatment = str_replace(replicate,"\\d",""))


ggplot(PAM,aes(x = date, y = fv_fm_y_1000)) + 
  
    geom_boxplot(aes(fill = treatment)) + labs(x = "Date", y = "Fv/Fm", title = "Fv/Fm by Date and Treatment") +
  theme_minimal()+
  stat_compare_means(aes(group = treatment), method = "anova", label = "p.format", size=3)

ggsave("../output/PAM.png", plot = last_plot(), width = 9, height = 4, bg = "white")

facet_PAM <- ggplot(PAM,aes(x = treatment, y = fv_fm_y_1000)) + 
    geom_boxplot(aes(fill = treatment)) + labs(x = "Date", y = "Fv/Fm", title = "Fv/Fm by Date and Treatment") +
    facet_grid(~timepoint )+
  theme_minimal()+
  stat_compare_means()

facet_PAM
  #stat_compare_means(aes(group = treatment), method = "anova", label = "p.format", size=3)

ggplot(PAM,aes(x = treatment, y = fv_fm_y_1000)) + 
    geom_boxplot(aes(fill = date)) + labs(x = "Treatment", y = "Fv/Fm", title = "Fv/Fm by Date and Treatment") +
  theme_minimal()+
  scale_fill_manual(values =c("20260421" = "orange", "20260423" = "olivedrab")) +
  stat_compare_means(aes(group = date), method = "anova", label = "p.format", size=3)

facet_PAM_treatment <- ggplot(PAM,aes(x = date, y = fv_fm_y_1000)) + 
     geom_boxplot(aes(color = date))+
    geom_line(aes(group = replicate)) + labs(x = "Date", y = "Fv/Fm", title = "Fv/Fm by Date and Treatment") +
    geom_point(aes(color = date), alpha = 0.5 ) +
    facet_grid(~treatment )+
  scale_color_manual(values =c("20260421" = "orange", "20260423" = "olivedrab")) +
  theme_minimal()+
  stat_compare_means(paired= TRUE)

facet_PAM_treatment




```



apex_Temp <- apex_Exp %>% 
  filter(ProbeType == "Tmp")

apex_Temp %>% 
  ggplot(aes(x = DateTimeEST, y = Value, color = ProbeName)) +
  geom_line(linewidth = 0.3, alpha = 0.4) +
  geom_smooth(se = FALSE, linewidth = 1) +
  facet_wrap(~ProbeName, ncol = 2) +
  labs(
    x = "Date Time",
    y = "Temperature (°C)",
    title = "Temperature Over Time"
  ) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "grey90"),
    legend.position = "none"
  )
```



#Make a list of dataframes, each containing a horizontal line that will correspond to the upper and lower threshold of each parameter (temperature, salinity, pH total)
hlines_data <- list(
  data.frame(yintercept = 22.0, metric = "Temp"), # lower threshold for temperature in C°
  data.frame(yintercept = 30.0, metric = "Temp"), # upper threshold for temperature in C°
  data.frame(yintercept = 33.0, metric = "Cond"), # lower threshold for salinity in psu
  data.frame(yintercept = 70.0, metric = "Cond"), # upper threshold for salinity in psu
  data.frame(yintercept = 7.5, metric = "pH"), # lower threshold for total pH
  data.frame(yintercept = 8.4, metric = "pH") # upper threshold for total pH
   )

#creating facet labels for the individual facets
facet_labels <- unique(apex_Exp$ProbeType)
names(facet_labels) = facet_labels
facet_labels <- replace(facet_labels, which(facet_labels == "pH"), "pH")
facet_labels <- replace(facet_labels, which(facet_labels == "Cond"), "Salinity (psu)")
facet_labels <- replace(facet_labels, which(facet_labels == "Temp"), "Temperature (ºC)")

apex_final<-apex_toplot %>%
  ggplot(aes(x=DateTimeEST, y=Value, colour=ProbeName))+
  geom_point(size=2)+
  xlab("DateTimeEST")+
  facet_grid(factor(ProbeType,c("pH","Cond","Temp")) ~ ., scales = "free_y", labeller = as_labeller(facet_labels))+
  geom_hline(data = hlines_data[[1]], aes(yintercept = yintercept), linetype = "dashed") +    
  geom_hline(data = hlines_data[[2]], aes(yintercept = yintercept), linetype = "dashed") +
  geom_hline(data = hlines_data[[3]], aes(yintercept = yintercept), linetype = "dashed") +
  geom_hline(data = hlines_data[[4]], aes(yintercept = yintercept), linetype = "dashed") +
  geom_hline(data = hlines_data[[5]], aes(yintercept = yintercept), linetype = "dashed") +
  geom_hline(data = hlines_data[[6]], aes(yintercept = yintercept), linetype = "dashed") +
  theme_bw() + 
  theme(text = element_text(size = 14)); apex_final


# Save plot 
ggsave("../Output/Daily_Measurements.pdf", apex_final, width = 20, height = 15, units = c("in"))

```


#exported apex data online
#1. Export data
#2.Open xml file in excel
#3. Save as csv
#4. Importing csv into R


#reading file
apex <- read.csv("C:/PUTNAM LAB/Pacuta_Polyp_Bailout/data/ApexLogger/Spring_2026/20260422Trail1/ApexRawData.csv", sep=",", skip=c(1), header=TRUE, na.strings = "NA")[6:9]

#Selecting column names of our desire
colnames(apex) <-  c("DateTime","ProbeName","ProbeType","Value")

head(apex)

#converting date and adapting to the timezone
apex$DateTimeHST <- parse_date_time(apex$DateTime, "%m/%d/%y %H:%M:%S", tz = "Pacific/Honolulu")
apex$DateTimeEST <- with_tz(apex$DateTimeHST, tzone = "America/New_York")
head(apex)

# filtering dates based on experiment timeline  (use GMT for the time, so +6)
apex_Exp <- apex %>% filter(DateTimeEST >= "2026-04-22 11:00:00" & DateTimeEST <= "2026-04-23 11:00:00")

1.00#filtering data to be plotted

apex_toplot <- apex_Exp %>%
  filter(ProbeType == "pH",
         ProbeType == "Cond",
         ProbeType == "Temp")

#apex_pH <- apex_Exp %>% filter (ProbeType== "pH")
#apex_Cond <- apex_Exp %>% filter (ProbeType == "Cond")
#apex_temps <- apex_Exp %>% filter(ProbeType == "Temp")

##Plot data 

#logger_plot %>% ggplot()



## plotting extracted data
#apex_temps %>% ggplot(aes(x=DateTimeEST, y=Value, color = ProbeName)) +
  #geom_line(aes(color = ProbeName), size = 0.5) +
  #facet_grid(~ProbeType)+
  #ylab("Temperature (°C)") +theme_minimal()


#ggsave("C:/PUTNAM LAB/Pacuta_Polyp_Bailout/output/20260422_apex_temps.pdf",
      #width = 10, height = 20, units = "in", , bg = "white")
``
